\documentclass{article} % For LaTeX2e
\usepackage{iclr2025_conference,times}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,amsfonts}  % Enhanced math support
\usepackage{booktabs}                   % Better tables
\usepackage{multirow}                   % Multi-row cells in tables
\usepackage{tabularx}                   % Tables with adjustable columns
\usepackage{xcolor}                     % Color support
\usepackage{microtype}                  % Better typography
\usepackage{subcaption}                 % Subfigures and subcaptions
\usepackage{algorithm2e}                % Algorithm formatting
\usepackage{tikz}                       % Drawing and diagrams
\usepackage{wrapfig}                    % Wrap text around figures
\usepackage{makecell}                   % Cell formatting with line breaks

% Optional math commands from https://github.com/goodfeli/dlbook_notation.
\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{url}
\usepackage{cleveref}                   % Better cross-referencing (must be after hyperref)
\usepackage{listings}                   % Code/prompt listings in boxes
\lstset{basicstyle=\ttfamily\small, frame=single, breaklines=true, columns=fullflexible, literate={<}{{\textless}}1 {>}{{\textgreater}}1}

% Define theorem-like environments
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]


\title{Can Models refuse questions they don't know? \\ Measuring Knowledge-aware Refusal in LLMs}

% Authors must not appear in the submitted version. They should be hidden
% as long as the \iclrfinalcopy macro remains commented out below.
% Non-anonymous submissions will be rejected without review.

\author{Antiquus S.~Hippocampus, Natalia Cerebro \& Amelie P. Amygdale \thanks{ Use footnote for providing further information
about author (webpage, alternative address)---\emph{not} for acknowledging
funding agencies.  Funding acknowledgements go at the end of the paper.} \\
Department of Computer Science\\
Cranberry-Lemon University\\
Pittsburgh, PA 15213, USA \\
\texttt{\{hippo,brain,jen\}@cs.cranberry-lemon.edu} \\
\And
Ji Q. Ren \& Yevgeny LeNet \\
Department of Computational Neuroscience \\
University of the Witwatersrand \\
Joburg, South Africa \\
\texttt{\{robot,net\}@wits.ac.za} \\
\AND
Coauthor \\
Affiliation \\
Address \\
\texttt{email}
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\cm}[1]{\textcolor{red}{#1}}
% \newcommand{\cm}[1]{}
\newcommand{\bt}[1]{\textcolor{blue}{#1}}

%\iclrfinalcopy % Uncomment for camera-ready version, but NOT for submission.
\begin{document}


\maketitle

\begin{abstract}
Large Language Models (LLMs) should refuse to answer questions beyond their knowledge, which is a crucial ability for LLMs to be reliable. 
However, existing metrics face three challenges in quantifying this ability: 
\textbf{(1) inaccurate:} yielding inconsistent and biased scores; 
\textbf{(2) complex:} requiring heavy calibration processes; 
and \textbf{(3) indirect:} measuring auxiliary models rather than refusal behavior. 
In this work, we propose a new metric called the \emph{Refusal Index (RI)} to measure how accurately LLMs refuse questions they do not know. 
We define RI as the Spearman's rank correlation between the refusal probability and the error probability. 
To calculate RI, we propose a simple and principled two-pass evaluation that estimates RI from refusal rates in two evaluation runs. 
Experiments across 16 models and 5 datasets show that RI quantifies a model's ability to make accurate knowledge-aware refusals. Unlike existing metrics, RI remains consistent when models exhibit different refusal rates.
RI addresses an important but previously untested aspect of LLM factual evaluation. 
Our observations reveal a blind spot in LLM factual evaluations: 
while LLMs are achieving higher accuracy on factual evaluations, they are not always improving their refusal behavior. 
This finding highlights the need to complement traditional metrics with the Refusal Index for comprehensive factual evaluation.

\end{abstract}

\input{Sections/introduction.tex}

\input{Sections/method.tex}

\input{Sections/experimental_setup.tex}

\input{Sections/results.tex}

\input{Sections/discussion.tex}

\input{Sections/related_work_conclusion.tex}

\bibliography{iclr2025_conference}
\bibliographystyle{iclr2025_conference}

\input{Sections/appendix.tex}

\end{document}